### 2021-07-09 rad-icfp-2021/relwork.tex
improving related work

An exception is \cite{abadi-plotkin2020} which gives a reverse-mode transformation that is proved correct and with the right complexity.
This work is purely first-order, does not have arrays and uses a non-standard semantics. 
Our work is more canonical in the sense that it can more directly be implemented in a standard call-by-value language.
They do allow conditionals and recursion but their semantics suggest that they don't have a canonical transformation on these constructs. 
Instead, they follow the principle of unrolling for recursive calls and evaluating the conditional before differentiating, 
which requires to recompute the whole gradient when taken at new values that can change the value of the conditional. 
This is a well-known limit in TensorFlow \cite{abadi2016tensorflow} which TeaserFlow EagerMode \cite{agrawal2019tensorflow} tries to bypass.

@@

Given the enormous computation needs for state-of-the-art large scale machine learning applications, 
which require extremely efficient tensor computations and automatic differentiation for backpropagation, 
combining array languages and automatic differentiation (tensor calculus) in the best 
fitted intermediate representation for optimizations is of key interest and active research.
\cite{bernstein2020differentiating} 
TODO
\cite{laue2018computing,laue2020simple}

@@

TODO
\cite{lee2020correctness}: non diff function
\cite{sherman2021}: non diff HO functions
\cite{pearlmutter2008reverse}: based on this
\cite{elliott2018simple}: bit similar because using CT as useful language, but we take impl and opti more seriously.
\cite{shaikhha2019efficient}: similar approach but here more complicated for reverse mode, would be interesting to investigate more loop fusions here.
\cite{paszke2021getting}
Curry paper: The idea of using closures as back-propagators is receiving recent attention. 
    For example Julia Zygote \cite{innes2019zygote} and Swift AD \cite{wei2018first} chose this design. 
    Other recent work follows similar ideas [27, 26] but is using meta-programming as an implementation technique.

Relatively recent work in AD (2008-2020):
\cite{mak2020differential,elliotthigher,vytiniotis2019differentiable,innes2018don,baydin2017automatic,huot2020correctness,gallagher-sdg,manzyuk2012confusion,lantern_icfp,beck1994if,wang2018backpropagation,betancourt2018geometric,elliott2018simple,carpenter2015stan,paszke2017automatic,shaikhha2019efficient,innes2019zygote,griewank2008evaluating,kucukelbir2017automatic,brunel2019backpropagation,barthe2020versatility,abadi2019simple,cockett2019reverse,van2018automatic,hascoet2013tapenade,abadi2016tensorflow,pearlmutter2008reverse,bergstra2010theano,fong2019backprop,ehrhard2003differential,agrawal2019tensorflow,bettencourt2019taylor,cruttwell2017cartesian,manzyuk2012simply,laue2018computing,bernstein2020differentiating}

Good recent surveys: \cite{van2018automatic,baydin2017automatic}

>>>>

% An exception is \cite{abadi-plotkin2020}.
% This work is purely first-order, does not have arrays and uses a non-standard semantics. 
% Our work is more canonical in the sense that it can more directly be implemented in a standard call-by-value language.
% They do allow conditionals and recursion but their semantics suggest that they don't have a canonical transformation on these constructs. 
% Instead, they follow the principle of unrolling the recursive calls and evaluating the conditional before differentiating, 
% which requires to recompute the whole gradient when taken at new values that can change the value of the conditional. 
% This is a known limitation in TensorFlow \cite{abadi2016tensorflow}. %which TeaserFlow EagerMode \cite{agrawal2019tensorflow} tries to bypass.

@@

Given the enormous computation needs for state-of-the-art large scale machine learning applications, 
which require extremely efficient tensor computations and automatic differentiation for backpropagation, 
combining array languages and automatic differentiation (tensor calculus) in the best-fitted 
intermediate representation for optimizations is of key interest and active research. 
See for instance the recent work by \cite{bernstein2020differentiating,laue2018computing,laue2020simple}.
Advanced array programming is considered an orthogonal problem to AD and we focused our work on the latter.

@@

Table~\ref{tbl:relwork} does not reflect all the aspects of AD. For instance \cite{lee2020correctness} studies in more detail non differentiability, 
and \cite{sherman2021} the differentiability of highly-non trivial higher-order functions. 
Our work is somewhat close in spirit to the idea of \cite{elliott2018simple} of compiling to categories.
The idea of using closures as back-propagators is receiving recent attention, as is highlighted in \cite{vytiniotis2019differentiable},
these ideas are used in Julia Zygote \cite{innes2019zygote}, Swift AD \cite{wei2018first}, and recently in \cite{paszke2021getting}.
These seem closer to using control mechanims than having purely functional reverse-derivatives as far as we could tell.
Other aspects of AD still are discussed in recent surveys \cite{van2018automatic,baydin2017automatic}.
